---
title: "Installation of R, RStudio and git"
output: 
  distill::distill_article:
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


**Please install the latest versions of `r fontawesome::fa_i("r-project")`,
RStudio, and `git`.**

`r fontawesome::fa_i("lightbulb")`
**Having an up-to-date environment is the first good practice.**


<br/>



## Installation



### `r fontawesome::fa_i("windows")` &nbsp;Windows



1. Download `r fontawesome::fa_i("r-project")` at:<br/>
[https://cran.r-project.org/](https://cran.r-project.org/)

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;`r fontawesome::fa_i("caret-right")` &nbsp;
Click on _Download R for Windows_
<br/>&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;`r fontawesome::fa_i("caret-right")` &nbsp;
Click on _base_
<br/>&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;`r fontawesome::fa_i("caret-right")` &nbsp;
Click on _Download R 4.1.2 for Windows_
<br/>&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;`r fontawesome::fa_i("caret-right")` &nbsp;
Open _R-4.1.2-win.exe_ and follow instructions



<br/>



2. Download **`RStudio Desktop`** at:<br/>
[https://www.rstudio.com/products/rstudio/download/#download](https://www.rstudio.com/products/rstudio/download/#download)

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;`r fontawesome::fa_i("caret-right")` &nbsp;
Download the version for Windows 10 (the **.exe** and not the **.zip**)
<br/>&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;`r fontawesome::fa_i("caret-right")` &nbsp;
Open _RStudio-2021.09.0-351.exe_ and follow instructions
<br/>&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;`r fontawesome::fa_i("caret-right")` &nbsp;
Check if **`RStudio`** has found `r fontawesome::fa_i("r-project")`. 
Launch **`RStudio`** and run:

```{r echo=TRUE, eval=FALSE}
R.version.string
## [1] "R version 4.1.2 (2021-11-01)"
```


<br/>



3. Download **`git`** at:<br/>
[https://git-scm.com/downloads](https://git-scm.com/downloads)

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;`r fontawesome::fa_i("caret-right")` &nbsp;
Download the 64-bit version for Windows
<br/>&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;`r fontawesome::fa_i("caret-right")` &nbsp;
Open _Git-2.33.1-64-bit.exe_ and follow instructions
<br/>&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;`r fontawesome::fa_i("caret-right")` &nbsp;
Check if **`RStudio`** has found **`git`**.
Launch **`RStudio`** and run:

```{r echo=TRUE, eval=FALSE}
Sys.which("git")
## C:/Program Files/Git/bin/git.exe
```


<br/>



4. Download **`Rtools4`** at:<br/>
[https://cran.r-project.org/bin/windows/Rtools/](https://cran.r-project.org/bin/windows/Rtools/)

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;`r fontawesome::fa_i("caret-right")` &nbsp;
Download the 64-bit version for Windows
<br/>&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;`r fontawesome::fa_i("caret-right")` &nbsp;
Open _rtools40v2-x86_64.exe_ and follow instructions

The installation of **`Rtools4`** is optional, but this software is required
to install packages (from sources) containing compiled code (C/C++/Fortran). 
**`Rtools4`** contains in particular the utility `make`.

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;`r fontawesome::fa_i("caret-right")` &nbsp;
Once **`Rtools4`** is installed, you need to tell `r fontawesome::fa_i("r-project")` 
where to find **`Rtools4`** utilities.
Open **`RStudio`** et run:


```{r echo=TRUE, eval=FALSE}
cat("PATH='${RTOOLS40_HOME}\\usr\\bin;${PATH}'\r\n", file = "~/.Renviron", append = TRUE)
```

An hidden file `.Renviron` has been created (or modified) in you personal folder
and contains the line `PATH='${RTOOLS40_HOME}\\usr\\bin;${PATH}'` (the path to **`Rtools4`**). 
At each start `r fontawesome::fa_i("r-project")` will read this file.

Restart **`RStudio`** and check that `r fontawesome::fa_i("r-project")` find `make`:

```{r echo=TRUE, eval=FALSE}
Sys.which("make")
## "C:/rtools40/usr/bin/make.exe"
```

Now you can install packages from sources:


```{r echo=TRUE, eval=FALSE}
install.packages("jsonlite", type = "source")
```


<br/>



5. Install **`pandoc`**

The package **`rmarkdown`** requires the software **`pandoc`** to convert document
from one format (e.g. `.Rmd`) to another (e.g. `.html`).

Download **`pandoc`** at:
[https://github.com/jgm/pandoc/releases/tag/2.16.1](https://github.com/jgm/pandoc/releases/tag/2.16.1),
and follow instructions.


<br/>



6. Install a **`LaTeX`** distribution

If you want to convert documents from `.Rmd` to `.pdf`, you will need **`LaTeX`**. 
You can install a lightened distribution using the `r fontawesome::fa_i("r-project")`
package [**`tinytex`**](https://yihui.org/tinytex/).
Open **`RStudio`** and run: 

```{r echo=TRUE, eval=FALSE}
## Install tinytex ----
install.packages("tinytex")

## Install LaTeX ----
tinytex::install_tinytex()
```

**N.B.** If you already have a **`LaTeX`** distribution, do not install `tinytex`.

<br/>



### `r fontawesome::fa_i("apple")` &nbsp;macOS


Th easiest way to set up your environment is by using the package manager for macOS
[**`Homebrew`**](https://brew.sh). With this tool you can install (and update) your
software directly from the terminal. First you need to install the Apple software
_Command Line Tools for Xcode_.

Open a [terminal](https://support.apple.com/en-gb/guide/terminal/apd5265185d-f365-44cb-8b09-71a064a42125/mac), 
run this line and accept the license:

```sh
## Install Xcode Command Line Tools ----
sudo xcode-select --install
```

Then you can install **`Homebrew`**:

```sh
## Install Homebrew ----
/bin/bash -c "$(curl -fsSL https://raw.githubusercontent.com/Homebrew/install/HEAD/install.sh)"

## Check installation ----
brew --version
## Homebrew 3.3.8

## Update repositories (list of available software) ----
brew update
```


Follow these lines to install your environment:

```sh
## Install utilities for R ----
brew install --cask xquartz     ## Graphical window system
brew install gfortran           ## FORTRAN compiler for GCC

## Install R ----
brew install --cask r

## Install RStudio Desktop ----
brew install --cask rstudio

## Install git ----
brew install git

## Install pandoc ----
brew install pandoc             ## Convert documents (Rmd, html, etc.)
```

Open **`RStudio`** and check your installation:

```{r echo=TRUE, eval=FALSE}
R.version.string
## "R version 4.1.2 (2021-11-01)"

Sys.which("git")
##            git
## "/usr/bin/git"

Sys.which("make")
##            make 
## "/usr/bin/make"

Sys.which("pandoc")
##            pandoc 
## "/usr/local/bin/pandoc"
```

<br/>

If you want to convert documents from `.Rmd` to `.pdf`, you will need **`LaTeX`**. 
You can install a lightened distribution using the `r fontawesome::fa_i("r-project")`
package [**`tinytex`**](https://yihui.org/tinytex/).
Open **`RStudio`** and run: 

```{r echo=TRUE, eval=FALSE}
## Install tinytex ----
install.packages("tinytex")

## Install LaTeX ----
tinytex::install_tinytex()
```


<br />


`r fontawesome::fa_i("lightbulb")` &nbsp;
Congrats! You have successfully install your environment. 
The following commands will install system libraries (software) required by
some `r fontawesome::fa_i("r-project")` packages (`sf`, `sp`, `rgdal`,
`rgeos`, `raster`, `terra`, `xml`, etc.).


```sh
## Install other utilities for R (spatial and others) ----
brew install libgit2 
brew install libxml2 
brew install poppler 
brew install geos gdal netcdf proj udunits
```

To keep your system up-to-date run frequently:


```sh
## Update environment ----
brew update && brew upgrade --greedy
```


<br/>



### `r fontawesome::fa_i("ubuntu")`&nbsp;Ubuntu



To keep your system up-to-date, do not forget to frequently run:

```sh
## System update ----
sudo apt update && sudo apt upgrade
```


<br/>


Then check your version of Ubuntu:


```sh
## Check Ubuntu version ----
cat /etc/issue
## Ubuntu 20.04 LTS
```

<br/>

| Release   | Release name   | Repository URL                                               |
|:---------:|:---------------|:-------------------------------------------------------------|
| 21.04     | Hirsute Hippo  | https://cloud.r-project.org/bin/linux/ubuntu hirsute-cran40/ |
| 20.10     | Groovy Gorilla | https://cloud.r-project.org/bin/linux/ubuntu groovy-cran40/  |
| 20.04 LTS | Focal Fossa    | https://cloud.r-project.org/bin/linux/ubuntu focal-cran40/   |
| 18.04 LTS | Bionic Beaver  | https://cloud.r-project.org/bin/linux/ubuntu bionic-cran40/  |
| 16.04 LTS | Xenial Xerus   | https://cloud.r-project.org/bin/linux/ubuntu xenial-cran40/  |
: Table 1. List of CRAN repositories for different versions of Ubuntu

<br/>

Install `r fontawesome::fa_i("r-project")` by following these steps 
(change the URL below by the one specific to your version of Ubuntu, cf. Table 1) :

```sh
## Install APT utilities ----
sudo apt install software-properties-common

## Add CRAN GPG key ----
sudo apt-key adv --keyserver keyserver.ubuntu.com --recv-keys E298A3A825C0D65DFD57CBB651716619E084DAB9

## Add CRAN repository to APT repositories list ----
sudo add-apt-repository 'deb https://cloud.r-project.org/bin/linux/ubuntu focal-cran40/'

## Update packages list ----
sudo apt update

## Install R ----
sudo apt install r-base

## Install R (for development, recommended) ----
sudo apt install build-essential
sudo apt install r-base-dev
```


<br/>


To install **`RStudio Desktop`** visit this page: 
https://www.rstudio.com/products/rstudio/download/#download, download the
`.deb` file for Ubuntu 18/Debian 10 et install it. If you have an older version 
of Ubuntu (16.04), visit this page: https://www.rstudio.com/products/rstudio/older-versions/

Alternatively, you can install **`RStudio Desktop`** from the terminal: 

```sh
## Install wget utility ----
sudo apt install wget

## Download RStudio Desktop installer (Ubuntu >= 18.04) ----
wget https://download1.rstudio.org/desktop/bionic/amd64/rstudio-2021.09.0-351-amd64.deb

## Download RStudio Desktop installer (Ubuntu 16.04 only) ----
# wget https://download1.rstudio.org/desktop/xenial/amd64/rstudio-1.4.1106-amd64.deb

## Install RStudio Desktop ----
sudo dpkg -i rstudio-2021.09.0-351-amd64.deb
# sudo dpkg -i rstudio-1.4.1106-amd64.deb

## Remove installer ----
rm rstudio-2021.09.0-351-amd64.deb
# rm rstudio-1.4.1106-amd64.deb
```

<br/>

Now let's install **`git`**: 

```sh
sudo apt install git
```

Finally, install **`pandoc`** (to convert documents using `rmarkdown`):

```sh
sudo apt install pandoc
```


<br/>


Open **`RStudio`** and check your installation:

```{r echo=TRUE, eval=FALSE}
R.version.string
## "R version 4.1.2 (2021-11-01)"

Sys.which("git")
##            git
## "/usr/bin/git"

Sys.which("make")
##            make 
## "/usr/bin/make"

Sys.which("pandoc")
##            pandoc 
## "/usr/bin/pandoc"
```

<br/>

If you want to convert documents from `.Rmd` to `.pdf`, you will need **`LaTeX`**. 
You can install a lightened distribution using the `r fontawesome::fa_i("r-project")`
package [**`tinytex`**](https://yihui.org/tinytex/).
Open **`RStudio`** and run: 

```{r echo=TRUE, eval=FALSE}
## Install tinytex ----
install.packages("tinytex")

## Install LaTeX ----
tinytex::install_tinytex()
```


<br />


`r fontawesome::fa_i("lightbulb")` &nbsp;
Congrats! You have successfully install your environment. 
The following commands will install system libraries (software) required by
some `r fontawesome::fa_i("r-project")` packages (`sf`, `sp`, `rgdal`,
`rgeos`, `raster`, `terra`, `xml`, etc.).


```sh
## Install other utilities for R (spatial and others) ----
sudo apt install libgdal-dev libproj-dev libgeos-dev libudunits2-dev libnetcdf-dev
sudo apt install libcairo2-dev libmagick++-dev
sudo apt install libpoppler-cpp-dev
```

To keep your system up-to-date, do not forget to frequently run:


```sh
## Update system ----
sudo apt update && sudo apt upgrade
```


<br/>



## Configuration


### `git`


To use **`git`**, you need to store your credentials (user name and email)
that will be added to all your commits.
Open a terminal (you can use the terminal of **RStudio**) and run:

```sh
git config --global user.name  "Jane Doe"
git config --global user.email "jane.doe@mail.com"
```

When you create a new git project the default branch will be named _master_. 
It's recommended to replace _master_ by _main_ (for more information read this
[article](https://www.zdnet.com/article/github-to-replace-master-with-alternative-term-to-avoid-slavery-references/)).

On a terminal, run:

```sh
git config --global init.defaultBranch "main"
```

From now when you will initialize `git` in a new project the default branch will be **main**.


<br/>


### `r fontawesome::fa_i("github")`&nbsp;GitHub SSH keys

To communicate with [GitHub](https://github.com), you can use the protocol HTTPS (default)
or the protocol SSH (recommended). The protocol SSH is a more secure method of 
logging into a remote server (GitHub). To use this protocol, you need to generate 
an SSH keys pair that consists in creating two long strings of characters: a **public key**
and a **private key**. You can place the public key on any server (for instance GitHub), 
and then connect to the server using a device (your laptop) that has access to the private key.


To generate a new SSH keys pair proceed as follow: 

&nbsp;&nbsp;`r fontawesome::fa_i("caret-right")` &nbsp;
Create an account on **GitHub** at [https://github.com/signup](https://github.com/signup)
<br/>&nbsp;&nbsp;`r fontawesome::fa_i("caret-right")` &nbsp;
Open **`RStudio`** and click on: `Tools > Global options > Git/SVN`
<br/>&nbsp;&nbsp;`r fontawesome::fa_i("caret-right")` &nbsp;
Click on **Create RSA Key** and then click on **Create** 
(you can omit the passphrase). This new SSH keys pair has been stored in `~/.ssh/`.
<br/>&nbsp;&nbsp;`r fontawesome::fa_i("caret-right")` &nbsp;
Click on **View public key** and copy it
<br/>&nbsp;&nbsp;`r fontawesome::fa_i("caret-right")` &nbsp;
Go to the page [https://github.com/settings/keys](https://github.com/settings/keys)
and click on **New SSH key**
<br/>&nbsp;&nbsp;`r fontawesome::fa_i("caret-right")` &nbsp;
Choose a name for your key (e.g. your laptop name) and paste your public key.
<br/>&nbsp;&nbsp;`r fontawesome::fa_i("caret-right")` &nbsp;
Finally click on **Add SSH key**

<br />

`r fontawesome::fa_i("ubuntu")` &nbsp; If you are on Ubuntu you need to install the
following utility: 

```sh
sudo apt install ssh-askpass
```


<br />

Let's test the SSH connection with GitHub. Open a terminal and run (you need to accept
the fingerprint, just enter `yes`):

```sh
ssh -T git@github.com
```

Congrats! You can now communicate with GitHub using the SSH protocol.

<br />

`r fontawesome::fa_i("exclamation-triangle")` &nbsp; From now, you need to
use the SSH URL when connecting a new GitHub project (named `remote`).


```sh
## URL HTTPS ----
https://github.com/frbcesab/datatoolbox.git

## URL SSH (use this one) ----
git@github.com:frbcesab/datatoolbox.git
```

<br />

`r fontawesome::fa_i("exclamation-triangle")` &nbsp; 
If you lose your private SSH key you won't be able to establish
a connection with GitHub. You'll need to delete your SSH key
on GitHub and to create a new one.


<br/>


### `r fontawesome::fa_i("github")`&nbsp;GitHub PAT

If you want to use the GitHub API, essentially with the `r fontawesome::fa_i("r-project")` package
[`usethis`](https://usethis.r-lib.org/),
you need to create a GitHub Personal Access Token (PAT).

Go to [https://github.com/settings/tokens](https://github.com/settings/tokens)

&nbsp;&nbsp;`r fontawesome::fa_i("caret-right")` &nbsp;
Click on **Generate new token**
<br/>&nbsp;&nbsp;`r fontawesome::fa_i("caret-right")` &nbsp;
Choose a name for your new token (e.g. your laptop name)
<br/>&nbsp;&nbsp;`r fontawesome::fa_i("caret-right")` &nbsp;
In the section _scopes_, select at least **repo** and **workflow**
<br/>&nbsp;&nbsp;`r fontawesome::fa_i("caret-right")` &nbsp;
Copy this token.

Then open **`RStudio`** and run: 

```{r echo = TRUE, eval = FALSE}
## Install usethis package ----
install.packages("usethis")

## Open ~/.Renviron file ----
usethis::edit_r_environ()
## • Modify '~/.Renviron'
## • Restart R for changes to take effect
```

In the `~/.Renviron` file, add this new line: `GITHUB_PAT='XXX'` where `XXX` is your token.
Do not forget to restart `r fontawesome::fa_i("r-project")`.

<br />

Finally let's tell `usethis` to use the SSH protocol. Open **RStudio** and run:

```{r echo = TRUE, eval = FALSE}
## Install usethis package ----
install.packages("usethis")

## Open ~/.Rprofile file ----
usethis::edit_r_profile()
## • Modify '~/.Rprofile'
## • Restart R for changes to take effect
```

In the `~/.Rprofile` file, add this new line: `options(usethis.protocol = "ssh")`.
Do not forget to restart `r fontawesome::fa_i("r-project")`.



## Corrections {.appendix}


If you see mistakes or want to suggest changes, please 
[Create an issue](https://github.com/rdatatoolbox/rdatatoolbox.github.io/issues)
on the source repository.



## Reuse {.appendix}



The material of this website is licensed under Creative Commons Attribution 
[CC BY 4.0](https://creativecommons.org/licenses/by/4.0/).
Source code is available at https://github.com/rdatatoolbox/rdatatoolbox.github.io/.


## Citation {.appendix}



Casajus N, Dray S, Gimenez O, Guéry L, Guilhaumon F & Schiettekatte NMD (2021) 
Workshop FRB-CESAB & GdR EcoStat: Data Toolbox for Reproducible Research in 
Ecology (Edition 2021). 
Zenodo. <http://doi.org/10.5281/zenodo.4262978>.
